---
title: "Statistical Power Analysis: Goanna Burrow Monitoring — BACI Design (NTA-KJ Pilot)"
date: today
format:
  html:
    toc: true
    toc-depth: 3
    number-sections: true
    code-fold: true
    theme: cosmo
    embed-resources: true
execute:
  echo: false
  warning: false
  message: false
---

```{r setup}
library(tidyverse)
library(knitr)

# All outputs are loaded from pre-computed RDS files (produced by
# src/goanna-burrow-power/00_load_pilot_data.R and 03_run_simulations.R, run
# from the project root).
baselines           <- readRDS("../out/baselines.rds")
encounter_rate_grid <- readRDS("../out/encounter_rate_grid.rds")
species_grid        <- readRDS("../out/species_grid.rds")

effect_labels <- c(
  reduction_25pct = "25% relative reduction",
  reduction_50pct = "50% relative reduction",
  reduction_75pct = "75% relative reduction"
)
sigma_labels <- c(
  low      = "Low inter-annual variability",
  moderate = "Moderate inter-annual variability",
  high     = "High inter-annual variability"
)

min_days_for_power <- function(grid, group_cols, power_threshold = 0.80) {
  grid |>
    group_by(across(all_of(group_cols))) |>
    arrange(days_per_cell) |>
    summarise(
      min_days_80 = {
        hit <- days_per_cell[power >= power_threshold]
        if (length(hit) > 0) min(hit) else NA_real_
      },
      max_days_tested = max(days_per_cell),
      .groups = "drop"
    )
}
```

## Overview

This document reports a simulation-based statistical power analysis for a Before–After
Control–Impact (BACI) monitoring design based on the `goanna-hunting-app` burrow-search protocol,
using the first field tranche of pilot data (NTA-KJ, 3 field days). It is a sibling analysis to
`power_analysis.qmd` (sand-plot/road-transect track detection), sharing the same overall BACI
structure — 3 survey years before an impact event and 3 after — but differing in three structural
respects that follow from how goanna-burrow searches are actually conducted:

- **No fixed sites.** Each burrow search is a one-off event at wherever the crew happens to be
  searching, not a revisit to a permanent plot. The unit of replication is therefore the individual
  search (for the encounter-rate response) or the individual inspected burrow (for the
  species-composition response), and inter-annual variation is modelled with a random intercept for
  **year** rather than for site.
- **Two different response variables and error distributions**, addressing two different management
  questions:
  1. Is the rate at which searches encounter a **wandidji** burrow declining? (exponential
     time-to-event)
  2. Among burrows that get found and inspected, is the proportion containing each of 4 species
     declining? (binomial, one model per species)
- **Sample size is a consequence of effort and encounter rate, not a free design input.** How many
  searches (or inspected burrows) a fixed budget of field days actually yields depends on how
  quickly burrows are found — so the design axis reported throughout is **field days of effort per
  design cell**, not a raw sample size.

## Data

Pilot data come from the NTA-KJ portion of the first `goanna-hunting-app` field tranche, read
directly from the `data-warehousing` research-export CSV (see
`../../goanna-burrow-analysis/CLAUDE.md` for the extraction pipeline).

```{r tbl-pilot-summary}
#| label: tbl-pilot-summary
#| tbl-cap: "NTA-KJ pilot summary."

tibble(
  Quantity = c(
    "Field days",
    "Total searches",
    "Wandidji burrow events (imputed)",
    "Wandidji burrow events (drop first day)",
    "Inspected burrows"
  ),
  Value = c(
    baselines$n_field_days,
    baselines$lambda0_impute$n_searches,
    round(baselines$lambda0_impute$wandidji_events, 1),
    round(baselines$lambda0_drop_day1$wandidji_events, 1),
    baselines$species_baselines$n_inspected[1]
  )
) |>
  kable(align = c("l", "r"))
```

### A field-protocol gap in the pilot

`burrow_type` (the field used to identify wandidji burrows) was not recorded at all on the first
field day, and is still partly missing on the second — the same "not always recorded early in the
trial" issue, not two separate problems. An untyped found burrow is not the same as "no burrow", so
it cannot simply be counted as a non-wandidji outcome. Two baseline wandidji encounter-rate ($\lambda_0$)
estimates are therefore carried through the whole analysis as a labelled sensitivity pair, rather
than picking one silently:

- **impute** — every untyped found burrow (any date) is counted as a *fractional* wandidji event,
  at the empirical wandidji rate among burrows where the type *was* recorded. This preserves the
  person-time-at-risk contributed by every search, including the untyped-day searches.
- **drop_day1** — the fully-untyped first field day is excluded outright (both from the event count
  and from the time-at-risk denominator); the smaller residual gap on day 2 is still imputed the
  same way (dropping day 1 doesn't change the empirical rate used, since day 1 contributed zero
  typed rows to begin with).

```{r tbl-lambda0-variants}
#| label: tbl-lambda0-variants
#| tbl-cap: "Baseline wandidji encounter-rate estimates under the two missing-data treatments."

tibble(
  Variant                 = c("Impute", "Drop first day"),
  `Wandidji events`       = round(c(baselines$lambda0_impute$wandidji_events,
                                     baselines$lambda0_drop_day1$wandidji_events), 1),
  `Searches`              = c(baselines$lambda0_impute$n_searches,
                               baselines$lambda0_drop_day1$n_searches),
  `Field days`            = c(baselines$lambda0_impute$n_field_days,
                               baselines$lambda0_drop_day1$n_field_days),
  `lambda0 (events/sec)`  = signif(c(baselines$lambda0_impute$lambda0,
                                      baselines$lambda0_drop_day1$lambda0), 3),
  `Throughput (searches/day)` = round(c(baselines$lambda0_impute$throughput_per_day,
                                         baselines$lambda0_drop_day1$throughput_per_day), 1)
) |>
  kable(align = c("l", "r", "r", "r", "r", "r"))
```

The two variants differ by only about
`r round(100 * abs(baselines$lambda0_impute$lambda0 - baselines$lambda0_drop_day1$lambda0) / baselines$lambda0_impute$lambda0, 0)`%
— small enough that, as the results below show, it makes little practical difference to the power
curves.

### Species baselines

Denominator = every inspected burrow, regardless of whether `burrow_type` was recorded (inspection
is a separate field, unaffected by the gap above). "Wanggali" and "wandidji" are Karajarri names for
sand goanna and spiny-tailed goanna respectively; king brown and northern bluetongue are recorded
directly under those codes.

```{r tbl-species-baselines}
#| label: tbl-species-baselines
#| tbl-cap: "Baseline presence probability among inspected burrows, by species."

baselines$species_baselines |>
  arrange(desc(p0)) |>
  transmute(
    Species        = species_label,
    `Field code`   = species_code,
    `N present`    = n_present,
    `N inspected`  = n_inspected,
    `Baseline p0`  = round(p0, 4)
  ) |>
  kable(align = c("l", "l", "r", "r", "r"))
```

## Methods

### Study design

Both responses share a 2 (control/impact) $\times$ 2 (before/after) BACI structure over 6 survey
years (3 before, 3 after). With no fixed sites, the year itself becomes the natural grouping for
inter-annual variation: a **random intercept for year**, crossed with treatment (both control and
impact areas are surveyed within the same years) and nested within period (each year belongs to
exactly one of before/after by construction). This is the same mechanism that makes BACI designs
work in general — a year with unusual conditions shifts both treatment arms together, so the
treatment $\times$ period contrast is estimated net of that shared variation rather than confounded
with it.

### Response 1: wandidji burrow encounter rate

Each search is treated as a time-to-event trial. Let $T$ be the time to encountering a wandidji
burrow and $\lambda$ the hazard rate:

$$
\log(\lambda_{ijt}) = \log(\lambda_0) + u_j + \log(\text{effect}) \cdot \text{After}_t \times \text{Impact}_i
$$

where $j$ indexes year (1–6), $u_j \sim \mathcal{N}(0, \sigma^2_{\text{year}})$, and `effect` is the
proportion of baseline rate retained in the impact/after cell. A search ending in anything other
than a wandidji burrow (a different burrow type found, or nothing found) is **right-censored** at
its observed duration rather than dropped — filtering the dataset down to wandidji-only outcomes
would hide a real decline in encounter rate behind an unchanged "time among successes" average,
since a real decline would show up as *more* non-wandidji outcomes, not just slower ones.

Fitted as a Poisson GLMM with a `log(duration)` offset — the standard, computationally cheap
equivalent of censored exponential regression — testing the period $\times$ treatment interaction
via a Wald $z$-test:

$$
\texttt{event} \sim \texttt{period * treatment} + (1 \mid \texttt{year}) + \texttt{offset(log(duration))}, \quad \text{family = Poisson}
$$

**Sample size is derived, not a free input.** The number of searches achievable in a design cell is

$$
n_{\text{cell}} = \texttt{days\_per\_cell} \times \texttt{throughput}_0 \times m_{\text{cell}}
$$

where $m_{\text{cell}} = 1$ in every cell except impact/after, where $m_{\text{cell}} = \text{effect}$
— the same rate decline that reduces the hazard also means fewer searches get completed in the same
days of fieldwork. The unit of replication remains the individual search; what changes with effort
and effect size is how many such replicates are obtained.

### Response 2: proportion of inspected burrows containing each species

For each of the 4 species independently (king brown, wanggali, wandidji, northern bluetongue), a
binomial GLMM with the same year random intercept:

$$
\text{logit}(p_{jt}) = \text{logit}(p_0) + u_j + \texttt{After}_t \times \texttt{Impact}_i \cdot \beta_3
$$

Two distinct effect parameters govern each simulated dataset, because "fewer burrows get found" and
"the mix of species inside found burrows changes" are different hypotheses that happen to share a
sample-size mechanism:

- **`activity_effect`** — the proportion of baseline *overall* (any-type) burrow-finding activity
  retained in the impact/after cell. This drives sample size only (via the same
  $n_{\text{cell}}$ formula as response 1, using the inspected-burrow throughput), not the tested
  coefficient. By default it is set equal to response 1's `effect`, on the simplifying assumption
  that overall goanna-burrow activity and wandidji-specific activity decline together — a documented
  assumption, not a measured fact.
- **`composition_effect`** — the proportion of baseline species-presence probability retained in the
  impact/after cell. This *is* the parameter tested via the period $\times$ treatment interaction.

In the results below, `composition_effect = activity_effect` throughout (the coupled default).

### Simulation parameters

- **Effect sizes**: the same three scenarios as `power_analysis.qmd` — 25%, 50%, and 75% relative
  reduction (i.e. 75%, 50%, and 25% of baseline retained).
- **Inter-annual variability ($\sigma_{\text{year}}$)**: unmeasured — the pilot is one month, so it
  carries no information about year-to-year variance. Swept as low/moderate/high scenarios rather
  than fixed at a guessed value; "moderate" is loosely anchored to the existing sand-plot analysis's
  own Goanna between-plot logit-SD (~1.15) as an order-of-magnitude reference only.
- **Field-effort grid (`days_per_cell`)**: response 1 uses 1–20 days/cell (its baseline per-search
  event probability is high enough that a handful of field days already gives strong power);
  response 2 uses 5–80 days/cell, an order of magnitude larger, since the rarest species
  (baseline prevalence below 1%) need much more effort before power curves become informative.
- 200 Monte Carlo replicates per grid cell, `parallel::mcmapply` across up to 6 cores, seed 9034.
- With only 6 year-levels, singular/boundary GLMM fits are expected, especially at low effort and
  for rare species — the fraction of singular fits is tracked for every grid cell (see
  @sec-caveats).

## Results

### Wandidji burrow encounter rate

```{r fig-encounter-rate-power}
#| label: fig-encounter-rate-power
#| fig-cap: "Simulated power to detect a decline in wandidji burrow encounter rate, vs field
#|   days of effort per design cell. Facets: inter-annual variability. Colour: minimum
#|   detectable effect. Linetype: baseline lambda0 treatment. Dashed line = 80% power."
#| fig-width: 10
#| fig-height: 5

encounter_rate_grid |>
  mutate(
    effect_label     = factor(effect_labels[effect_label], levels = effect_labels),
    sigma_year_label = factor(sigma_labels[sigma_year_label], levels = sigma_labels),
    lambda0_variant  = str_to_title(str_replace(lambda0_variant, "_", " "))
  ) |>
  ggplot(aes(x = days_per_cell, y = power, colour = effect_label,
             linetype = lambda0_variant, group = interaction(effect_label, lambda0_variant))) +
  geom_hline(yintercept = 0.80, linetype = "dashed", colour = "grey40") +
  geom_point(size = 1.8) +
  geom_line() +
  facet_wrap(~sigma_year_label) +
  scale_y_continuous(labels = scales::percent_format(), limits = c(0, 1)) +
  scale_colour_manual(
    values = c("25% relative reduction" = "#E41A1C",
               "50% relative reduction" = "#377EB8",
               "75% relative reduction" = "orange"),
    name = "Minimum detectable effect"
  ) +
  labs(x = "Field days of effort per design cell", y = "Simulated power",
       linetype = "Baseline lambda0 estimate") +
  theme_bw() +
  theme(
    legend.position  = "bottom",
    legend.box       = "vertical",
    strip.background = element_rect(fill = "grey92"),
    panel.grid.minor = element_blank()
  )
```

```{r tbl-min-days-encounter}
#| label: tbl-min-days-encounter
#| tbl-cap: "Minimum field days per design cell at which simulated power first reaches 80%,
#|   by effect size, inter-annual variability, and lambda0 treatment. NA = 80% power not
#|   reached within the tested range (up to 20 days/cell)."

min_days_for_power(encounter_rate_grid, c("lambda0_variant", "effect_label", "sigma_year_label")) |>
  mutate(
    lambda0_variant  = str_to_title(str_replace(lambda0_variant, "_", " ")),
    effect_label     = effect_labels[effect_label],
    sigma_year_label = sigma_labels[sigma_year_label]
  ) |>
  arrange(effect_label, sigma_year_label, lambda0_variant) |>
  rename(`Lambda0 treatment` = lambda0_variant, `Effect` = effect_label,
         `Inter-annual variability` = sigma_year_label,
         `Min. days for 80% power` = min_days_80, `Max. days tested` = max_days_tested) |>
  kable(align = c("l", "l", "l", "r", "r"))
```

Power rises quickly with field days for this response. Under moderate inter-annual variability, a
75% or 50% relative reduction reaches 80% power within about 5–10 days of effort per cell; a 25%
reduction is markedly harder to detect — power still falls short of 80% even at 20 days/cell across
all three variability scenarios, meaning subtle declines in wandidji encounter rate would need more
effort than is explored here to detect reliably. The two $\lambda_0$ treatments (impute vs.
drop first day) track closely throughout, confirming that the missing-`burrow_type` handling
doesn't materially change the conclusions.

### Proportion of inspected burrows containing each species

```{r fig-species-power}
#| label: fig-species-power
#| fig-cap: "Simulated power to detect a decline in the proportion of inspected burrows
#|   containing each species, vs field days of effort per design cell. Facets: species (rows)
#|   x inter-annual variability (columns). Colour: minimum detectable effect
#|   (composition_effect = activity_effect). Dashed line = 80% power."
#| fig-width: 10
#| fig-height: 9

species_grid |>
  mutate(
    effect_label     = factor(effect_labels[effect_label], levels = effect_labels),
    sigma_year_label = factor(sigma_labels[sigma_year_label], levels = sigma_labels)
  ) |>
  ggplot(aes(x = days_per_cell, y = power, colour = effect_label, group = effect_label)) +
  geom_hline(yintercept = 0.80, linetype = "dashed", colour = "grey40") +
  geom_point(size = 1.8) +
  geom_line() +
  facet_grid(species_label ~ sigma_year_label) +
  scale_y_continuous(labels = scales::percent_format(), limits = c(0, 1)) +
  scale_colour_manual(
    values = c("25% relative reduction" = "#E41A1C",
               "50% relative reduction" = "#377EB8",
               "75% relative reduction" = "orange"),
    name = "Minimum detectable effect"
  ) +
  labs(x = "Field days of effort per design cell", y = "Simulated power") +
  theme_bw() +
  theme(
    legend.position  = "bottom",
    strip.background = element_rect(fill = "grey92"),
    panel.grid.minor = element_blank()
  )
```

```{r tbl-min-days-species}
#| label: tbl-min-days-species
#| tbl-cap: "Minimum field days per design cell at which simulated power first reaches 80%,
#|   by species, effect size, and inter-annual variability. NA = 80% power not reached within
#|   the tested range (up to 80 days/cell)."

min_days_for_power(species_grid, c("species_label", "effect_label", "sigma_year_label")) |>
  mutate(
    effect_label     = effect_labels[effect_label],
    sigma_year_label = sigma_labels[sigma_year_label]
  ) |>
  arrange(species_label, effect_label, sigma_year_label) |>
  rename(Species = species_label, Effect = effect_label,
         `Inter-annual variability` = sigma_year_label,
         `Min. days for 80% power` = min_days_80, `Max. days tested` = max_days_tested) |>
  kable(align = c("l", "l", "l", "r", "r"))
```

Detecting a decline in species composition needs an order of magnitude more effort than the
encounter-rate response, tracking each species' baseline prevalence (@tbl-species-baselines):
wandidji (the most common, `r round(100*baselines$species_baselines$p0[baselines$species_baselines$species_label=="wandidji"],1)`%)
and king brown (`r round(100*baselines$species_baselines$p0[baselines$species_baselines$species_label=="king_brown"],1)`%)
reach 80% power for a 75% reduction at around 40 days/cell; northern bluetongue
(`r round(100*baselines$species_baselines$p0[baselines$species_baselines$species_label=="northern_bluetongue"],1)`%)
needs closer to 80 days/cell for the same. Wanggali — the rarest species at only
`r round(100*baselines$species_baselines$p0[baselines$species_baselines$species_label=="wanggali"],2)`%
baseline prevalence — does not reach 80% power anywhere in the tested range (up to 80 days/cell,
any effect size or variability scenario): a decline in wanggali occupancy of burrows would likely
not be reliably detectable at feasible monitoring effort.

## Caveats {#sec-caveats}

```{r tbl-singular-fits}
#| label: tbl-singular-fits
#| tbl-cap: "Grid cells with more than 20% singular/boundary GLMM fits (unreliable year-variance
#|   estimation), by response. A high singular fraction means the power estimate for that cell
#|   should be read with caution rather than taken at face value."

bind_rows(
  encounter_rate_grid |>
    transmute(response = paste0("encounter_rate (", lambda0_variant, ")"),
              days_per_cell, effect_label = effect_labels[effect_label],
              sigma_year_label = sigma_labels[sigma_year_label], singular_fraction),
  species_grid |>
    transmute(response = species_label, days_per_cell,
              effect_label = effect_labels[effect_label],
              sigma_year_label = sigma_labels[sigma_year_label], singular_fraction)
) |>
  filter(singular_fraction > 0.20) |>
  arrange(desc(singular_fraction)) |>
  mutate(singular_fraction = round(singular_fraction, 2)) |>
  rename(Response = response, `Days/cell` = days_per_cell, Effect = effect_label,
         `Inter-annual variability` = sigma_year_label,
         `Singular fit fraction` = singular_fraction) |>
  kable(align = c("l", "r", "l", "l", "r"))
```

Several considerations bear on how these results should be used:

- **Pilot-scale baselines.** All baseline rates/proportions come from one month of data at a single
  native title area (139 searches, 3 field days) — a trial run intended to test process, not a
  mature baseline. Point estimates, especially for the rarer species (as few as 1–5 detections),
  carry wide uncertainty that the $\sigma_{\text{year}}$ and effect-size sensitivity sweeps only
  partly capture.
- **Singular fits.** With only 6 year-levels, the year-variance component is estimated imprecisely,
  and many grid cells (concentrated at low effort and for rare species — up to 82% of fits in the
  worst cells) return a singular/boundary fit. Power estimates in those cells should be treated as
  approximate.
- **Coupled activity/composition effects.** Response 2's sample size assumes overall burrow-finding
  activity and species-specific composition decline together (`activity_effect = composition_effect`
  throughout). If these were to diverge in practice, the actual achievable sample size — and hence
  power — could differ from what's shown here.

## Software and reproducibility

All analyses were conducted in R (v`r getRversion()`). Key packages: `lme4` v`r
packageVersion("lme4")` for GLMM fitting, `tidyverse` v`r packageVersion("tidyverse")` for data
manipulation and visualisation. Scripts are in `src/goanna-burrow-power/`; intermediate results are
cached in `out/` as `.rds` files (`baselines.rds`, `encounter_rate_grid.rds`, `species_grid.rds`).
