---
title: "Statistical Power Analysis: Sand-Plot Track Detection — BACI Design"
date: today
format:
  html:
    toc: true
    toc-depth: 3
    number-sections: true
    code-fold: true
    theme: cosmo
    embed-resources: true
execute:
  echo: false
  warning: false
  message: false
---

```{r setup}
library(tidyverse)
library(knitr)

# All outputs are loaded from pre-computed RDS files (produced by
# src/03_run_simulations.R, run from the project root).
power_full  <- readRDS("../out/power_full_grid.rds")
power_focal <- readRDS("../out/power_focal_grid.rds")
species_df  <- readRDS("../out/species_baselines.rds")

effect_labels <- c(
  reduction_25pct = "25% relative reduction",
  reduction_50pct = "50% relative reduction",
  reduction_75pct = "75% relative reduction"
)
```

## Overview

This document reports a simulation-based statistical power analysis for a
Before–After Control–Impact (BACI) monitoring design in which animal track
detections at sand plots are the primary response variable. The goal is to
determine the number of plots required to reliably detect a biologically
meaningful decline in track detection probability following an impact event.

## Methods

### Study design

The proposed design follows a two-treatment ($\times$ two-period) BACI
structure:

- **Treatments**: impact plots (subject to disturbance) and control plots
  (undisturbed reference sites).
- **Periods**: three survey years before the impact event and three years after.
- Each plot is surveyed once per year under standardised conditions (same
  season, cleared substrate, narrow weather window), so the primary observation
  at each plot–visit is binary: tracks of a given species detected (1) or not
  (0).

The parameter of interest is the **time × treatment interaction**: a departure
from parallel temporal trends between impact and control groups that is
attributable to the impact.

### Statistical model

Each simulated dataset is analysed with a binomial generalised linear mixed
model (GLMM):

$$
\text{logit}(p_{ijt}) = \beta_0 + \beta_1 \cdot \text{After}_t
  + \beta_2 \cdot \text{Impact}_i
  + \beta_3 \cdot \text{After}_t \times \text{Impact}_i
  + u_i
$$

where $i$ indexes plots, $t$ indexes survey years,
$u_i \sim \mathcal{N}(0, \sigma^2_{\text{plot}})$ is a plot-level random
intercept, and $\beta_3$ is the BACI interaction coefficient. Power is
estimated as the proportion of simulations in which the Wald $z$-test for
$\beta_3$ rejects $H_0 : \beta_3 = 0$ at $\alpha = 0.05$.

Models were fitted with the `lme4` package [@bates2015] using the BOBYQA
optimiser.

### Simulation parameters

**Baseline detection probabilities** were derived from historical sand-plot
survey data (Table 9 of the project baseline report). 

**Between-plot variability** ($\sigma_{\text{plot}}$) was estimated from a binomial GLM with random effects of site, from previous data on repeat-surveyed sites. 

**Minimum detectable effect (MDE)**: three scenarios are compared:

- **25% relative reduction** — detection probability in impact plots falls to
  75% of its pre-impact baseline while controls remain stable (e.g.
  $p_0 = 0.40 \rightarrow p_{\text{impact, after}} = 0.30$).
- **50% relative reduction** — detection probability halves
  (e.g. $p_0 = 0.40 \rightarrow p_{\text{impact, after}} = 0.20$).
- **75% relative reduction** — detection probability drops by three quarters
  (e.g. $p_0 = 0.40 \rightarrow p_{\text{impact, after}} = 0.10$).


**Sample size range**: $n = 20$–$50$ plots per treatment group (the proposed
field budget), with an extended grid of $n = 20$–$200$ evaluated for four
representative focal species.

All simulations used 200 Monte Carlo replicates per cell. Parallel computation
used `parallel::mcmapply` across up to six CPU cores; the random seed was fixed
at 9034 for reproducibility.

## Baseline detection probabilities

```{r tbl-baselines}
#| label: tbl-baselines
#| tbl-cap: "Baseline detection probabilities for all species recorded in
#|   historical sand-plot surveys (n = 460 plots). Species are ordered by
#|   detection frequency."

species_df |>
  filter(p_baseline < 1) |>
  select(species, detections, p_baseline, sigma) |>
  arrange(desc(p_baseline)) |> 
  mutate(
    p_baseline = round(p_baseline, 3),
    detections = as.integer(detections)
  ) |>
  rename(
    Species    = species,
    Detections = detections,
    `Baseline p` = p_baseline,
    Sigma = sigma
  ) |>
  kable(align = c("l", "r", "r"))
```

## Power results

### Power within the proposed plot budget (n = 20–50)

@fig-power-vs-p shows simulated power across all species for plots per group
in the range 20–50, for all three MDEs. The 80% power threshold (dashed line) is
not approached by any species or sample size under the 25% MDE. Under the 50%
MDE, power remains well below 80% for all but the most commonly detected
species even at $n = 50$. Under the 75% MDE, power approaches or exceeds 80%
for the most detectable species at $n = 50$.

```{r fig-power-vs-p}
#| label: fig-power-vs-p
#| fig-cap: "Simulated power vs baseline detection probability for three minimum
#|   detectable effects (facets) and three plot counts per group (colours).
#|   Points are individual species; curves are loess smooths.
#|   Dashed line = 80% power."
#| fig-width: 12
#| fig-height: 4.5

power_full |>
  filter(p_baseline < 1, p_baseline >= 0.01) |>
  mutate(
    n_label      = factor(paste0("n = ", n_per_group),
                          levels = paste0("n = ", c(20, 35, 50))),
    effect_label = factor(effect_labels[effect_label], levels = effect_labels)
  ) |>
  ggplot(aes(x = p_baseline, y = power, colour = n_label, group = n_label)) +
  geom_hline(yintercept = 0.80, linetype = "dashed", colour = "grey40") +
  geom_point(size = 1.8, alpha = 0.8) +
  geom_smooth(se = FALSE, span = 1.2, linewidth = 0.8) +
  annotate("text", x = 0.54, y = 0.84, label = "80% power",
           colour = "grey35", size = 3.2, hjust = 1) +
  facet_wrap(~effect_label, ncol = 3) +
  scale_x_continuous(labels = scales::percent_format(), limits = c(0, 0.60)) +
  scale_y_continuous(labels = scales::percent_format(), limits = c(0, 1)) +
  scale_colour_brewer(palette = "Set1", name = "Plots per group") +
  labs(x = "Baseline detection probability", y = "Simulated power") +
  theme_bw() +
  theme(
    legend.position  = "bottom",
    strip.background = element_rect(fill = "grey92"),
    panel.grid.minor = element_blank()
  )
```

### Summary table at n = 35 plots per group

```{r tbl-power-n35}
#| label: tbl-power-n35
#| tbl-cap: "Simulated power at n = 35 plots per treatment group for all
#|   species with baseline detection probability ≥ 1%. All three MDEs are shown.
#|   Monte Carlo SE ≈ 0.024 for each cell (200 replicates)."

power_full |>
  filter(p_baseline < 1, p_baseline >= 0.01, n_per_group == 35) |>
  select(species, p_baseline, effect_label, power) |>
  pivot_wider(names_from = effect_label, values_from = power) |>
  arrange(desc(p_baseline)) |>
  mutate(
    p_baseline        = round(p_baseline, 3),
    reduction_25pct   = round(reduction_25pct, 2),
    reduction_50pct   = round(reduction_50pct, 2),
    reduction_75pct   = round(reduction_75pct, 2)
  ) |>
  rename(
    Species           = species,
    `Baseline p`      = p_baseline,
    `Power (25% MDE)` = reduction_25pct,
    `Power (50% MDE)` = reduction_50pct,
    `Power (75% MDE)` = reduction_75pct
  ) |>
  kable(align = c("l", "r", "r", "r", "r"))
```

### Power curves for focal species

@fig-power-curves extends the sample size range to $n = 200$ plots per group
for four species representing a spread of baseline detection probabilities.
These illustrate the approximate plot count required to achieve 80% power.

```{r fig-power-curves}
#| label: fig-power-curves
#| fig-cap: "Simulated power vs plots per treatment group for four focal
#|   species, under all three MDEs. Facet labels show species common names.
#|   Dashed line = 80% power."
#| fig-width: 8
#| fig-height: 6

power_focal |>
  mutate(
    effect_label = factor(effect_labels[effect_label], levels = effect_labels),
    species      = str_remove(species, " \\(.*\\)")
  ) |>
  ggplot(aes(x = n_per_group, y = power,
             colour = effect_label, group = effect_label)) +
  geom_hline(yintercept = 0.80, linetype = "dashed", colour = "grey40") +
  geom_point(size = 2) +
  geom_line() +
  annotate("text", x = 198, y = 0.84, label = "80% power",
           colour = "grey35", size = 3.2, hjust = 1) +
  facet_wrap(~species, nrow = 2) +
  scale_x_continuous(breaks = c(20, 50, 100, 150, 200)) +
  scale_y_continuous(labels = scales::percent_format(), limits = c(0, 1)) +
  scale_colour_manual(
    values = c("25% relative reduction" = "#E41A1C",
               "50% relative reduction" = "#377EB8",
               "75% relative reduction" = "orange"),
    name   = "Minimum detectable effect"
  ) +
  labs(x = "Plots per treatment group", y = "Simulated power") +
  theme_bw() +
  theme(
    legend.position  = "bottom",
    strip.background = element_rect(fill = "grey92"),
    panel.grid.minor = element_blank()
  )
```

## Discussion

### Key findings

Under the proposed field budget of 20–50 plots per treatment group, power is
low for most species regardless of which MDE is assumed (@fig-power-vs-p,
@tbl-power-n35):

- **25% relative reduction**: power does not exceed ~25% for any species.
  The design cannot reliably distinguish a 25% decline from sampling noise.
- **50% relative reduction**: power reaches ~40–60% for the most detectable
  species (e.g. *Felis catus*, *Macrotis lagotis*) at $n = 50$, but still
  falls well short of the conventional 80% threshold.
- **75% relative reduction**: power approaches or exceeds 80% at $n = 50$ for
  species with the highest baseline detection probabilities, suggesting that
  very large impacts are detectable within the proposed plot budget.

Achieving 80% power for moderate effects requires substantially larger plot
counts than are currently budgeted (@fig-power-curves). For a 50% MDE,
approximately 100–150 plots per group are needed for species with baseline
$p \geq 0.35$. For a 25% MDE, sample requirements exceed 200 plots per group
for all species.

### Practical recommendations

1. **Target species selection**: Focus power calculations on the handful of
   focal species with highest baseline $p$ (e.g. *Felis catus*,
   *Macrotis lagotis*, *Varanus* sp.), where even moderate sample sizes
   provide some discrimination.
2. **Increase n or accept a larger MDE**: If the budget cannot support
   $n \geq 100$ plots per group, the study should be framed as capable of
   detecting only large effects (≥ 50–75% reduction) and reported with
   explicit power caveats.
3. **Reduce residual variance**: Stratified placement of plots by habitat
   type, or including habitat covariates in the model, may reduce
   $\sigma_{\text{plot}}$ and improve power without adding plots.
4. **Multiple species composite**: A multi-species index (e.g. species
   richness of native fauna tracks) may have lower variance than any single
   species and could yield higher power for a given n.

## Software and reproducibility

All analyses were conducted in R (v4.4.2). Key packages: `lme4` v`r
packageVersion("lme4")` for GLMM fitting, `tidyverse` v`r
packageVersion("tidyverse")` for data manipulation and visualisation. Scripts
are in `src/`; intermediate results are cached in `out/` as `.rds` files.

## References

::: {#refs}
:::
